---
title: "Comparing DESeq2 and edgeR with VISTA"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    fig_width: 8
    fig_height: 6
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  # BiocStyle crops figures via magick; skip it so a runner
  # without magick does not emit a warning per figure.
  crop = NULL,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

ensure_pkgs_available <- function(pkgs) {
  missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) {
    stop("Missing required packages: ", paste(missing, collapse = ", "))
  }
}

# This vignette requires all dependencies to already be installed.
ensure_pkgs_available(c("VISTA", "ggplot2", "dplyr", "patchwork"))
```

# Introduction

DESeq2 and edgeR are the two most widely used tools for differential expression
(DE) analysis of RNA-seq data. Both use negative binomial models but differ in
normalization, dispersion estimation, and statistical testing, which can lead to
different sets of differentially expressed genes (DEGs).

This vignette demonstrates how to use **VISTA** to run both methods on the same
dataset and systematically compare their results using built-in visualization
functions.

## What you will learn

- How to create VISTA objects with DESeq2 and edgeR backends
- How to compare DEG counts, fold-change estimates, and significance calls
- How to identify concordant and discordant genes between methods
- How to visualize method agreement with volcano plots, scatter plots, Venn
  diagrams, and heatmaps

# Setup

```{r load-packages}
library(VISTA)
library(ggplot2)
library(dplyr)
library(patchwork)

# Set a global theme with larger fonts for all ggplot2 plots
theme_set(theme_minimal(base_size = 16))
```

# Data

We use the built-in VISTA example data derived from the **airway** dataset
(Himes et al. 2014), which contains RNA-seq counts from human airway smooth
muscle cells treated with dexamethasone.

```{r load-data}
data("count_data", package = "VISTA")
data("sample_metadata", package = "VISTA")

dim(count_data)
head(sample_metadata[, c("sample_names", "cond_long")])
```

# Create VISTA Objects

We create two VISTA objects from the same data --- one using DESeq2 and one
using edgeR --- with identical cutoffs so that any differences in results are
attributable to the method.

```{r create-deseq2}
vista_deseq <- create_vista(
  counts        = count_data,
  sample_info   = sample_metadata,
  column_geneid = "gene_id",
  group_column  = "dex",
  group_numerator   = "trt",
  group_denominator = "untrt",
  method        = "deseq2",
  log2fc_cutoff = 1,
  pval_cutoff   = 0.05,
  min_counts    = 10,
  min_replicates = 1
)
vista_deseq
```

```{r create-edger}
vista_edger <- create_vista(
  counts        = count_data,
  sample_info   = sample_metadata,
  column_geneid = "gene_id",
  group_column  = "dex",
  group_numerator   = "trt",
  group_denominator = "untrt",
  method        = "edger",
  log2fc_cutoff = 1,
  pval_cutoff   = 0.05,
  min_counts    = 10,
  min_replicates = 1
)
vista_edger
```

Both objects contain the same comparison (treated vs untreated):

```{r comparison-names}
names(comparisons(vista_deseq))
names(comparisons(vista_edger))
```

# DEG Summary Comparison

## DEG counts per method

A quick way to compare the methods is to count the number of up- and
down-regulated genes detected by each.

```{r deg-barplots, fig.width=12, fig.height=6}
p_deseq <- get_deg_count_barplot(vista_deseq) +
  ggtitle("DESeq2") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_edger <- get_deg_count_barplot(vista_edger) +
  ggtitle("edgeR") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_deseq + p_edger +
  plot_annotation(title = "Number of DEGs by Method")
```

## DEG composition (pie and donut)

```{r deg-circular, fig.width=12, fig.height=8}
pie_deseq <- get_deg_count_pieplot(vista_deseq, facet_by = "none") +
  ggtitle("DESeq2 (Pie)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pie_edger <- get_deg_count_pieplot(vista_edger, facet_by = "none") +
  ggtitle("edgeR (Pie)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

donut_deseq <- get_deg_count_donutplot(vista_deseq, facet_by = "none") +
  ggtitle("DESeq2 (Donut)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

donut_edger <- get_deg_count_donutplot(vista_edger, facet_by = "none") +
  ggtitle("edgeR (Donut)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

(pie_deseq + pie_edger) / (donut_deseq + donut_edger)
```

## Tabular summary

```{r deg-table}
summarise_degs <- function(vista, method_name) {
  comps <- comparisons(vista)
  do.call(rbind, lapply(names(comps), function(comp) {
    de <- as.data.frame(comps[[comp]])
    n_up   <- sum(de$regulation == "Up", na.rm = TRUE)
    n_down <- sum(de$regulation == "Down", na.rm = TRUE)
    data.frame(
      Method     = method_name,
      Comparison = comp,
      Up         = n_up,
      Down       = n_down,
      Total_DEG  = n_up + n_down,
      Tested     = nrow(de)
    )
  }))
}

deg_summary_table <- rbind(
  summarise_degs(vista_deseq, "DESeq2"),
  summarise_degs(vista_edger, "edgeR")
)

knitr::kable(deg_summary_table, row.names = FALSE,
             caption = "DEG counts by method and comparison")
```

# Volcano Plots

Volcano plots show the relationship between fold-change magnitude and
statistical significance. Comparing them side-by-side reveals differences in
how each method distributes p-values and fold-change estimates.

```{r volcano-plots, fig.width=14, fig.height=7}
comp1 <- names(comparisons(vista_deseq))[1]

v_deseq <- get_volcano_plot(vista_deseq, sample_comparison = comp1) +
  ggtitle(paste("DESeq2 -", comp1))

v_edger <- get_volcano_plot(vista_edger, sample_comparison = comp1) +
  ggtitle(paste("edgeR -", comp1))

v_deseq + v_edger
```

# MA Plots

MA plots display fold-change as a function of mean expression. They are useful
for assessing whether low-expression genes show inflated fold-changes and how
each method handles shrinkage.

```{r ma-plots, fig.width=14, fig.height=7}
ma_deseq <- get_ma_plot(vista_deseq, sample_comparison = comp1) +
  ggtitle(paste("DESeq2 -", comp1))

ma_edger <- get_ma_plot(vista_edger, sample_comparison = comp1) +
  ggtitle(paste("edgeR -", comp1))

ma_deseq + ma_edger
```

# Fold-Change Concordance

## Log2 fold-change scatter plot

A direct comparison of fold-change estimates between methods reveals their
overall concordance. Genes on the diagonal have identical estimates; deviations
indicate method-specific differences.

```{r fc-scatter, fig.width=7, fig.height=6}
# Extract DE results for the first comparison
de_deseq <- as.data.frame(comparisons(vista_deseq)[[comp1]])
de_edger <- as.data.frame(comparisons(vista_edger)[[comp1]])

# Merge on gene_id
merged <- merge(
  de_deseq[, c("gene_id", "log2fc", "padj", "regulation")],
  de_edger[, c("gene_id", "log2fc", "padj", "regulation")],
  by = "gene_id",
  suffixes = c("_deseq2", "_edger")
)

# Concordance category
merged$concordance <- case_when(
  merged$regulation_deseq2 == "Up" & merged$regulation_edger == "Up"     ~ "Both Up",
  merged$regulation_deseq2 == "Down" & merged$regulation_edger == "Down" ~ "Both Down",
  merged$regulation_deseq2 == "Other" & merged$regulation_edger == "Other" ~ "Not DE",
  TRUE ~ "Discordant"
)

conc_colors <- c(
  "Both Up"    = "#D73027",
  "Both Down"  = "#4575B4",
  "Discordant" = "#FFC107",
  "Not DE"     = "grey80"
)

# Correlation
r_val <- cor(merged$log2fc_deseq2, merged$log2fc_edger,
             use = "complete.obs", method = "pearson")

ggplot(merged, aes(x = log2fc_deseq2, y = log2fc_edger, color = concordance)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = conc_colors) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
           label = paste0("Pearson r = ", round(r_val, 3)),
           fontface = "italic", size = 4) +
  labs(
    x = "log2 Fold Change (DESeq2)",
    y = "log2 Fold Change (edgeR)",
    color = "Concordance",
    title = paste("Fold-Change Agreement:", comp1)
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## Concordance summary table

```{r concordance-table}
conc_table <- merged %>%
  count(concordance) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(desc(n))

knitr::kable(conc_table, col.names = c("Category", "Genes", "Percent"),
             caption = "Concordance between DESeq2 and edgeR calls")
```

# P-value Comparison

## P-value scatter plot

Comparing adjusted p-values (on the -log10 scale) shows whether both methods
agree on which genes are most significant.

```{r pvalue-scatter, fig.width=7, fig.height=6}
merged$neg_log10_padj_deseq2 <- -log10(merged$padj_deseq2 + 1e-300)
merged$neg_log10_padj_edger  <- -log10(merged$padj_edger + 1e-300)

ggplot(merged, aes(x = neg_log10_padj_deseq2, y = neg_log10_padj_edger,
                   color = concordance)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = conc_colors) +
  labs(
    x = "-log10(padj) DESeq2",
    y = "-log10(padj) edgeR",
    color = "Concordance",
    title = paste("Adjusted P-value Agreement:", comp1)
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## P-value distribution

The distribution of raw p-values provides a diagnostic for each method. A
well-calibrated test produces a uniform distribution under the null, with a
spike near zero for truly DE genes.

```{r pvalue-histogram, fig.width=10, fig.height=4}
pval_df <- rbind(
  data.frame(pvalue = de_deseq$pvalue, Method = "DESeq2"),
  data.frame(pvalue = de_edger$pvalue, Method = "edgeR")
)

ggplot(pval_df, aes(x = pvalue, fill = Method)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~Method) +
  scale_fill_manual(values = c("DESeq2" = "#1B9E77", "edgeR" = "#D95F02")) +
  labs(
    x = "Raw P-value",
    y = "Count",
    title = "P-value Distribution by Method"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

# Venn Diagram of DEGs

VISTA provides `get_deg_venn_diagram()` for comparing DEG overlap across
comparisons within a single object. For cross-method comparison (DESeq2 vs
edgeR), we use `get_genes_by_regulation()` to extract gene lists from each
VISTA object, then pass them to `ggvenn`.

## Upregulated genes

```{r venn-up, fig.width=7, fig.height=5}
# Extract upregulated gene sets using VISTA's accessor
up_deseq <- get_genes_by_regulation(
  vista_deseq, sample_comparisons = comp1, regulation = "Up"
)[[1]]

up_edger <- get_genes_by_regulation(
  vista_edger, sample_comparisons = comp1, regulation = "Up"
)[[1]]

venn_up <- list(DESeq2 = up_deseq, edgeR = up_edger)
ggvenn::ggvenn(venn_up, fill_color = c("#1B9E77", "#D95F02")) +
  ggtitle(paste("Upregulated Genes:", comp1))
```

## Downregulated genes

```{r venn-down, fig.width=7, fig.height=5}
down_deseq <- get_genes_by_regulation(
  vista_deseq, sample_comparisons = comp1, regulation = "Down"
)[[1]]

down_edger <- get_genes_by_regulation(
  vista_edger, sample_comparisons = comp1, regulation = "Down"
)[[1]]

venn_down <- list(DESeq2 = down_deseq, edgeR = down_edger)
ggvenn::ggvenn(venn_down, fill_color = c("#1B9E77", "#D95F02")) +
  ggtitle(paste("Downregulated Genes:", comp1))
```

## All DEGs (Up + Down)

```{r venn-all, fig.width=7, fig.height=5}
all_deseq <- union(up_deseq, down_deseq)
all_edger <- union(up_edger, down_edger)
venn_all  <- list(DESeq2 = all_deseq, edgeR = all_edger)

ggvenn::ggvenn(venn_all, fill_color = c("#1B9E77", "#D95F02")) +
  ggtitle(paste("All DEGs:", comp1))
```

# Normalization Comparison

DESeq2 uses median-of-ratios normalization while edgeR uses TMM
(trimmed mean of M-values). Comparing the resulting normalized count
distributions helps assess whether normalization choices drive
downstream differences.

```{r normalization-density, fig.width=10, fig.height=5}
nc_deseq <- norm_counts(vista_deseq)
nc_edger <- norm_counts(vista_edger)

# Use common genes
common_genes <- intersect(rownames(nc_deseq), rownames(nc_edger))
sample_cols  <- intersect(colnames(nc_deseq), colnames(nc_edger))

norm_df <- rbind(
  data.frame(
    log2_count = as.vector(log2(nc_deseq[common_genes, sample_cols] + 1)),
    Method = "DESeq2"
  ),
  data.frame(
    log2_count = as.vector(log2(nc_edger[common_genes, sample_cols] + 1)),
    Method = "edgeR"
  )
)

ggplot(norm_df, aes(x = log2_count, fill = Method)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("DESeq2" = "#1B9E77", "edgeR" = "#D95F02")) +
  labs(
    x = "log2(Normalized Count + 1)",
    y = "Density",
    title = "Normalized Count Distribution"
  ) +
  theme_minimal(base_size = 16)
```

## Sample-level comparison

```{r norm-scatter-samples, fig.width=10, fig.height=8}
# Compare normalized counts for each sample
sample_norm_df <- data.frame(
  DESeq2 = as.vector(log2(nc_deseq[common_genes, sample_cols] + 1)),
  edgeR  = as.vector(log2(nc_edger[common_genes, sample_cols] + 1)),
  Sample = rep(sample_cols, each = length(common_genes))
)

ggplot(sample_norm_df, aes(x = DESeq2, y = edgeR)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~Sample, ncol = 3) +
  labs(
    x = "log2(DESeq2 Normalized + 1)",
    y = "log2(edgeR Normalized + 1)",
    title = "Per-Sample Normalized Counts"
  ) +
  theme_minimal(base_size = 16)
```

# PCA Comparison

Principal component analysis reveals whether the two normalization
strategies preserve the same sample-level structure.

```{r pca-comparison, fig.width=14, fig.height=7}
pca_deseq <- get_pca_plot(vista_deseq) +
  ggtitle("PCA: DESeq2 Normalization")

pca_edger <- get_pca_plot(vista_edger) +
  ggtitle("PCA: edgeR Normalization")

pca_deseq + pca_edger
```

# Expression of Discordant Genes

Genes called as DEG by one method but not the other are of particular
interest. Visualizing their expression patterns helps assess whether
method-specific calls are biologically meaningful.

```{r discordant-genes}
deseq_only <- setdiff(all_deseq, all_edger)
edger_only <- setdiff(all_edger, all_deseq)

cat("DEGs unique to DESeq2:", length(deseq_only), "\n")
cat("DEGs unique to edgeR:", length(edger_only), "\n")
cat("DEGs shared:", length(intersect(all_deseq, all_edger)), "\n")
```

## Expression of DESeq2-only genes

```{r deseq-only-boxplot, fig.width=8, fig.height=5}
if (length(deseq_only) >= 3) {
  # Show top genes by fold change
  deseq_only_de <- de_deseq[de_deseq$gene_id %in% deseq_only, ]
  top_deseq_only <- head(
    deseq_only_de[order(-abs(deseq_only_de$log2fc)), "gene_id"],
    6
  )
  get_expression_boxplot(vista_deseq, genes = top_deseq_only) +
    ggtitle("Top DESeq2-Only DEGs (by |log2FC|)")
}
```

## Expression of edgeR-only genes

```{r edger-only-boxplot, fig.width=8, fig.height=5}
if (length(edger_only) >= 3) {
  edger_only_de <- de_edger[de_edger$gene_id %in% edger_only, ]
  top_edger_only <- head(
    edger_only_de[order(-abs(edger_only_de$log2fc)), "gene_id"],
    6
  )
  get_expression_boxplot(vista_edger, genes = top_edger_only) +
    ggtitle("Top edgeR-Only DEGs (by |log2FC|)")
}
```

# Sensitivity Analysis: Effect of Cutoffs

Different significance and fold-change thresholds can shift the balance between
methods. Here we sweep across a range of cutoffs.

```{r cutoff-sweep, fig.width=9, fig.height=5}
sweep_cutoffs <- function(de_deseq, de_edger, fc_vals, pval_cut = 0.05) {
  results <- lapply(fc_vals, function(fc) {
    n_deseq <- sum(
      abs(de_deseq$log2fc) >= fc & de_deseq$padj < pval_cut,
      na.rm = TRUE
    )
    n_edger <- sum(
      abs(de_edger$log2fc) >= fc & de_edger$padj < pval_cut,
      na.rm = TRUE
    )
    data.frame(
      log2fc_cutoff = fc,
      DESeq2 = n_deseq,
      edgeR  = n_edger
    )
  })
  do.call(rbind, results)
}

fc_vals <- seq(0.5, 3, by = 0.25)
sweep_df <- sweep_cutoffs(de_deseq, de_edger, fc_vals)

sweep_long <- tidyr::pivot_longer(
  sweep_df,
  cols = c("DESeq2", "edgeR"),
  names_to = "Method",
  values_to = "n_DEGs"
)

ggplot(sweep_long, aes(x = log2fc_cutoff, y = n_DEGs, color = Method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("DESeq2" = "#1B9E77", "edgeR" = "#D95F02")) +
  labs(
    x = "|log2FC| Cutoff",
    y = "Number of DEGs",
    title = "DEG Sensitivity to Fold-Change Threshold",
    subtitle = paste("Adjusted p-value <", 0.05)
  ) +
  theme_minimal(base_size = 16)
```

# Correlation Heatmaps

Sample-sample correlation heatmaps can reveal whether both normalization
approaches preserve the same correlation structure.

```{r corr-heatmaps, fig.width=14, fig.height=7}
corr_deseq <- get_corr_heatmap(vista_deseq) +
  ggtitle("DESeq2: Sample Correlation")

corr_edger <- get_corr_heatmap(vista_edger) +
  ggtitle("edgeR: Sample Correlation")

corr_deseq + corr_edger
```

# Summary and Recommendations

```{r summary-metrics}
overlap    <- length(intersect(all_deseq, all_edger))
union_size <- length(union(all_deseq, all_edger))
jaccard    <- round(overlap / union_size, 3)

cat("━━━ Method Comparison Summary ━━━\n")
cat("Comparison:", comp1, "\n\n")
cat("DESeq2 DEGs:", length(all_deseq), "\n")
cat("edgeR DEGs:", length(all_edger), "\n")
cat("Shared DEGs:", overlap, "\n")
cat("DESeq2-only:", length(deseq_only), "\n")
cat("edgeR-only:", length(edger_only), "\n")
cat("Jaccard index:", jaccard, "\n")
cat("Fold-change Pearson r:", round(r_val, 3), "\n")
```

## Key takeaways

1. **Fold-change estimates are highly correlated** between DESeq2 and edgeR, as
   both methods model similar biological effects.

2. **Significance calls differ at the margins**. Genes near the decision
   boundary (p-value or fold-change cutoff) are more likely to be called by one
   method but not the other.

3. **Core DEG sets overlap substantially**. Genes with large fold-changes and
   low p-values are consistently identified by both methods.

4. **Method-specific DEGs tend to have borderline statistics**. Inspecting
   discordant genes (unique to one method) often reveals marginal significance
   in the other method as well.

5. **For robust results**, consider requiring DEGs to be called by both methods,
   or use the intersection as a high-confidence set and the union as a
   discovery set.

# Session Info

```{r session-info}
sessionInfo()
```
