---
title: "Preparing Counts and Metadata for VISTA"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    fig_width: 8
    fig_height: 6
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

ensure_pkgs_available <- function(pkgs) {
  missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) {
    stop("Missing required packages: ", paste(missing, collapse = ", "))
  }
}

ensure_pkgs_available(c("VISTA", "dplyr", "tibble"))
```

# Overview

VISTA expects two main inputs for `create_vista()`:

1. A count table with one gene identifier column plus one column per sample.
2. A sample metadata table (`sample_info`) containing `sample_names` and the
   grouping variables used in the analysis.

In practice, users often start from count tables whose column names are file
paths, alignment outputs, or sequencing-derived sample identifiers. Metadata can
also be incomplete or missing entirely. VISTA's input preparation helpers are
designed to standardize these objects before differential analysis starts:

- `read_vista_counts()`
- `derive_vista_metadata()`
- `read_vista_metadata()`
- `match_vista_inputs()`

# Load Example Data

```{r load-data}
library(VISTA)
library(dplyr)
library(tibble)

data("count_data", package = "VISTA")
data("sample_metadata", package = "VISTA")

dim(count_data)
head(sample_metadata[, c("sample_names", "cond_long")])
```

# 1. Standardize Counts

For an ordinary in-memory count table, `read_vista_counts()` keeps the existing
API contract and returns a `create_vista()`-ready structure.

```{r counts-basic}
prepared_counts <- read_vista_counts(
  count_data[seq_len(50), ],
  format = "matrix",
  gene_id_column = "gene_id"
)

names(prepared_counts)
head(prepared_counts$counts[, seq_len(4)])
head(prepared_counts$row_data)
```

The returned object includes:

- `counts`: standardized count table with `gene_id`
- `row_data`: feature metadata aligned to the counts
- `sample_names`: standardized sample columns
- `sample_name_map`: mapping from original to repaired sample names

# 2. Repair File-Derived Sample Names

RNA-seq count tables often inherit sample columns from alignment or quantification
files. VISTA can repair these names conservatively when they are clearly
file-derived and the repaired names remain unique.

```{r sample-name-repair}
counts_paths <- count_data[seq_len(20), c("gene_id", sample_metadata$sample_names[seq_len(4)]), drop = FALSE]

colnames(counts_paths)[2:5] <- c(
  "/proj/run/03_alignment/HET_1_U/HET_1_U_star_alignAligned.sortedByCoord.out.bam",
  "/proj/run/03_alignment/HET_1_ovary/HET_1_ovary_star_alignAligned.sortedByCoord.out.bam",
  "/proj/run/03_alignment/WT_1_U/WT_1_U_star_alignAligned.sortedByCoord.out.bam",
  "/proj/run/03_alignment/WT_1_ovary/WT_1_ovary_star_alignAligned.sortedByCoord.out.bam"
)

prepared_paths <- read_vista_counts(
  counts_paths,
  format = "matrix",
  gene_id_column = "gene_id",
  repair_sample_names = "auto"
)

prepared_paths$sample_name_map
```

Automatic repair currently handles patterns such as:

- full file paths
- STAR outputs (`Aligned.sortedByCoord.out.bam`, `ReadsPerGene.out.tab`)
- RSEM outputs (`.genes.results`, `.isoforms.results`)
- generic quantification files such as `quant.sf`
- lane/read suffixes like `_S1_L001_R1_001`

# 3. Derive Starter Metadata from Sample Names

If you do not yet have a metadata sheet, `derive_vista_metadata()` can create a
starter `sample_info` table from the count sample names.

## Split-based parsing

```{r derive-split}
derived_split <- derive_vista_metadata(
  prepared_paths,
  parser = "split",
  split = "_",
  fields = c("genotype", "replicate", "tissue")
)

derived_split
```

## Regex-based parsing

For sequencing accessions or other structured labels, regex parsing gives more
control:

```{r derive-regex}
derived_regex <- derive_vista_metadata(
  counts = NULL,
  sample_names = c("SRR1039508", "SRR1039509", "SRR1039512"),
  parser = "regex",
  pattern = "SRR(\\d+)",
  fields = "run_id"
)

derived_regex
```

## Template mode

Template mode adds empty placeholders for columns users commonly fill in next,
such as `group` and `batch`.

```{r derive-template}
derive_vista_metadata(
  counts = NULL,
  sample_names = c("sampleA", "sampleB"),
  parser = "none",
  return_type = "template"
)
```

# 4. Standardize Existing Metadata

If a metadata table already exists, `read_vista_metadata()` standardizes it into
the form expected by `create_vista()`.

```{r metadata-basic}
prepared_samples <- read_vista_metadata(sample_metadata)

head(prepared_samples[, c("sample_names", "cond_long")])
```

VISTA will infer `sample_names` from:

- an existing `sample_names` column
- non-default rownames
- common aliases such as `sample`, `sample_id`, or `Run`

# 5. Match Counts and Metadata

`match_vista_inputs()` aligns count columns and metadata rows so they can be
passed directly into `create_vista()`.

```{r matching}
matched_inputs <- match_vista_inputs(prepared_counts, prepared_samples)

matched_inputs$column_geneid
identical(matched_inputs$sample_info$sample_names, colnames(matched_inputs$counts)[-1])
```

When appropriate, you can drop unmatched samples instead of erroring:

```{r matching-drop}
sample_extra <- prepared_samples[seq_len(4), , drop = FALSE]
extra_row <- sample_extra[1, , drop = FALSE]
extra_row$sample_names <- "extra_sample"
if ("cond_long" %in% colnames(extra_row)) extra_row$cond_long <- "control"
if ("dex" %in% colnames(extra_row)) extra_row$dex <- "untrt"
if ("cell" %in% colnames(extra_row)) extra_row$cell <- "N999"
rownames(extra_row) <- "extra_sample"
sample_extra <- rbind(sample_extra, extra_row)

matched_drop <- match_vista_inputs(
  prepared_counts,
  sample_extra,
  drop_unmatched = TRUE
)

matched_drop$report
```

# 6. Create a VISTA Object

Once inputs are standardized, `create_vista()` is unchanged.

```{r create-vista}
vista <- create_vista(
  counts = matched_inputs$counts,
  sample_info = matched_inputs$sample_info,
  column_geneid = matched_inputs$column_geneid,
  group_column = "cond_long",
  group_numerator = "treatment1",
  group_denominator = "control",
  method = "deseq2",
  min_counts = 5,
  min_replicates = 1
)

vista
```

# Practical Guidance

- Use `read_vista_counts()` first whenever count columns come from external
  tools or file names.
- Use `derive_vista_metadata()` when you only have sample names and need a
  reviewable starter metadata sheet.
- Use `read_vista_metadata()` when you already have a metadata table and want
  VISTA to standardize it.
- Use `match_vista_inputs()` before `create_vista()` to catch alignment issues
  early.

# Session Info

```{r session-info}
sessionInfo()
```
