---
title: "Color and Palette Design with VISTA"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    fig_width: 8
    fig_height: 6
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Color and Palette Design with VISTA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

ensure_pkgs_available <- function(pkgs) {
  missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) {
    stop("Missing required packages: ", paste(missing, collapse = ", "))
  }
}

# This vignette requires all dependencies to already be installed.
ensure_pkgs_available(c("VISTA", "ggplot2", "dplyr", "tibble", "colorspace"))
```

# Overview

A common challenge in RNA-seq analysis is visual consistency: the same biological
group should keep the same color across PCA, MDS, expression, and summary plots.
When this is not controlled, interpretation becomes harder and figure legends can
become confusing.

This vignette focuses on VISTA's color system and shows how to:

1. Discover all built-in palette options.
2. Keep colors consistent across different plot families.
3. Change global color style in one place.
4. Apply manual group/comparison color maps when required.

The floating table of contents on the side provides a quick-scroll panel for
navigating sections.

# Load Packages and Data

```{r load-packages}
library(VISTA)
library(ggplot2)
library(dplyr)
library(tibble)
library(colorspace)

data("count_data", package = "VISTA")
data("sample_metadata", package = "VISTA")

# Keep vignette runtime moderate while preserving signal
count_small <- count_data[1:1500, ]

# Use all available samples
table(sample_metadata$cond_long)
```

# Build a Baseline VISTA Object

```{r create-vista-baseline}
vista_dark <- create_vista(
  counts = count_small,
  sample_info = sample_metadata,
  column_geneid = "gene_id",
  group_column = "cond_long",
  group_numerator = "treatment1",
  group_denominator = "control",
  group_palette = "Dark 2",
  comparison_palette = "Dark 3",
  min_counts = 5,
  min_replicates = 1
)

vista_dark
```

Inspect the stored palette metadata:

```{r inspect-color-metadata}
group_palette(vista_dark)
group_colors(vista_dark)
```

# Built-In Palette Choices

VISTA group palettes are based on colorspace qualitative HCL palettes.

```{r palette-choices}
palette_table <- colorspace::hcl_palettes(type = "qualitative") |>
  as.data.frame() |>
  tibble::rownames_to_column("palette") |>
  dplyr::select(palette)

n_palettes <- nrow(palette_table)

cat("Number of built-in qualitative palettes available in VISTA:", n_palettes, "\n\n")
palette_table
```

Create a visual palette atlas:

```{r palette-atlas, fig.width=10, fig.height=6}
palette_grid <- do.call(
  rbind,
  lapply(palette_table$palette, function(pal) {
    cols <- colorspace::qualitative_hcl(8, palette = pal)
    data.frame(
      palette = pal,
      idx = seq_along(cols),
      col = cols,
      stringsAsFactors = FALSE
    )
  })
)

palette_grid$palette <- factor(palette_grid$palette, levels = rev(palette_table$palette))

ggplot(palette_grid, aes(x = idx, y = palette, fill = col)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_identity() +
  scale_x_continuous(breaks = 1:8) +
  labs(
    x = "Color index within palette",
    y = NULL,
    title = "VISTA Palette Atlas (Qualitative HCL)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

# Consistency Across Plot Families

Use the same `VISTA` object and verify that group colors are preserved across
multiple visualization types.

```{r prep-genes}
comp_name <- names(comparisons(vista_dark))[1]
up_genes <- get_genes_by_regulation(
  vista_dark,
  sample_comparisons = comp_name,
  regulation = "Up"
)[[comp_name]]

if (length(up_genes) < 6) {
  genes_demo <- head(rownames(vista_dark), 6)
} else {
  genes_demo <- head(up_genes, 6)
}

genes_demo
```

## PCA (group colors from object metadata)

```{r pca-consistent, fig.width=8, fig.height=6}
get_pca_plot(
  vista_dark,
  label = TRUE,
  shape_by = "cell",
  point_size = 5
)
```

## MDS (same group colors)

```{r mds-consistent, fig.width=8, fig.height=6}
get_mds_plot(
  vista_dark,
  label = TRUE,
  shape_by = "cell",
  point_size = 5
)
```

## Expression barplot (same group colors)

```{r expr-bar-consistent, fig.width=9, fig.height=6}
get_expression_barplot(
  vista_dark,
  genes = genes_demo[1:4],
  log_transform = TRUE
)
```

## Expression boxplot (same group colors)

```{r expr-box-consistent, fig.width=10, fig.height=6}
get_expression_boxplot(
  vista_dark,
  genes = genes_demo[1:4],
  log_transform = TRUE,
  pool_genes = FALSE,
  by = "gene",
  facet_by = "none",
  fill_by = "group"
)
```

# Switching Global Style in One Argument

Rebuild with a different global palette. No plot-specific color edits are needed.

```{r create-vista-warm}
vista_warm <- create_vista(
  counts = count_small,
  sample_info = sample_metadata,
  column_geneid = "gene_id",
  group_column = "cond_long",
  group_numerator = "treatment1",
  group_denominator = "control",
  group_palette = "Warm",
  comparison_palette = "Set 3",
  min_counts = 5,
  min_replicates = 1
)

group_palette(vista_warm)
group_colors(vista_warm)
```

```{r pca-palette-compare, fig.width=12, fig.height=5}
p_dark <- get_pca_plot(vista_dark, label = FALSE, point_size = 5) +
  ggtitle("Dark 2")

p_warm <- get_pca_plot(vista_warm, label = FALSE, point_size = 5) +
  ggtitle("Warm")

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_dark, p_warm, ncol = 2)
} else {
  p_dark
  p_warm
}
```

# Comparison-Level Colors

For multi-comparison workflows, VISTA also stores comparison colors.

```{r create-vista-multi}
# Build two directed comparisons to demonstrate comparison color mapping
vista_multi <- create_vista(
  counts = count_small,
  sample_info = sample_metadata,
  column_geneid = "gene_id",
  group_column = "cond_long",
  group_numerator = c("treatment1", "control"),
  group_denominator = c("control", "treatment1"),
  group_palette = "Dark 2",
  comparison_palette = "Set 3",
  min_counts = 5,
  min_replicates = 1
)

S4Vectors::metadata(vista_multi)$comparison$colors
```

```{r fc-bar-comparison-colors, fig.width=10, fig.height=6}
get_foldchange_barplot(
  vista_multi,
  genes = head(rownames(vista_multi), 6),
  facet_by = "none"
)
```

# Manual Color Maps (Advanced)

If your lab or journal has fixed brand colors, you can set explicit named
vectors using built-in VISTA setters.

Available manual setters:

- `set_vista_group_colors(vista, color_map)`
- `set_vista_comparison_colors(vista, color_map)`

```{r apply-manual-colors}
custom_group_cols <- c(
  control = "#264653",
  treatment1 = "#E76F51"
)

custom_comp_cols <- c(
  treatment1_VS_control = "#6C5CE7",
  control_VS_treatment1 = "#00A896"
)

vista_custom <- set_vista_group_colors(vista_dark, custom_group_cols)
vista_multi_custom <- set_vista_comparison_colors(vista_multi, custom_comp_cols)

group_colors(vista_custom)
S4Vectors::metadata(vista_multi_custom)$comparison$colors
```

```{r custom-color-plots, fig.width=12, fig.height=5}
p_custom_group <- get_pca_plot(
  vista_custom,
  label = FALSE,
  point_size = 5
) + ggtitle("Custom group colors")

p_custom_comp <- get_foldchange_barplot(
  vista_multi_custom,
  genes = head(rownames(vista_multi_custom), 6),
  facet_by = "none"
) + ggtitle("Custom comparison colors")

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_custom_group, p_custom_comp, ncol = 2)
} else {
  p_custom_group
  p_custom_comp
}
```

# Practical Guidance

- Use `group_palette` and `comparison_palette` in `create_vista()` for fast, global styling.
- Use `group_colors(vista)` to audit mapping before generating publication panels.
- Keep one palette per manuscript/project to preserve interpretation consistency.
- Switch to manual named vectors only when you need strict journal or brand colors.

# Session Information

```{r session-info}
sessionInfo()
```
