---
title: "Cell-Type Deconvolution with VISTA (airway)"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_width: 9
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Cell-Type Deconvolution with VISTA (airway)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

required_pkgs <- c(
  "VISTA",
  "airway",
  "SummarizedExperiment",
  "dplyr",
  "tibble",
  "tidyr",
  "ggplot2",
  "org.Hs.eg.db"
)
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing required packages: ", paste(missing_pkgs, collapse = ", "))
}

has_xcell2 <- requireNamespace("xCell2", quietly = TRUE)
has_deconv <- FALSE
```

# Overview

This vignette demonstrates how to run **cell-type deconvolution** with VISTA using the Bioconductor `airway` dataset.

Workflow covered:

1.  Build a `VISTA` object from airway counts and metadata.
2.  Run `run_cell_deconvolution()` (xCell2 backend).
3.  Extract cell-fraction estimates with `get_cell_fractions()`.
4.  Visualize sample-level composition with `plot_celltype_barplot()`.
5.  Compare treatment groups with `get_celltype_group_dotplot()`.
6.  Inspect cell-type/sample structure with `get_celltype_heatmap()`.

# Load Packages

```{r load-packages}
library(VISTA)
library(airway)
library(SummarizedExperiment)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(org.Hs.eg.db)
```

# Prepare Airway Data

```{r prepare-airway}
data("airway", package = "airway")

counts_matrix <- SummarizedExperiment::assay(airway, "counts")
sample_metadata <- as.data.frame(SummarizedExperiment::colData(airway))

# Keep sample IDs explicit for VISTA alignment.
sample_metadata$sample_names <- rownames(sample_metadata)
sample_metadata$cond_long <- ifelse(sample_metadata$dex == "trt", "dexamethasone", "untreated")

count_data <- as.data.frame(counts_matrix) %>%
  tibble::rownames_to_column("gene_id")

# Ensure column order follows sample_metadata.
count_data <- count_data[, c("gene_id", sample_metadata$sample_names)]

dim(count_data)
sample_metadata[, c("sample_names", "cell", "dex", "cond_long")]
```

# Create VISTA Object

```{r create-vista}
vista_airway <- create_vista(
  counts = count_data,
  sample_info = sample_metadata,
  column_geneid = "gene_id",
  group_column = "cond_long",
  group_numerator = "dexamethasone",
  group_denominator = "untreated",
  covariates = "cell",
  method = "deseq2",
  min_counts = 10,
  min_replicates = 1
)

# Add gene annotations (used for downstream interpretation and fallback symbol mapping).
vista_airway <- set_rowdata(
  vista_airway,
  orgdb = org.Hs.eg.db,
  columns = c("SYMBOL", "GENENAME", "ENTREZID"),
  keytype = "ENSEMBL"
)

vista_airway
```

# Run Deconvolution

`run_cell_deconvolution()` currently uses xCell2.\
If `xCell2` is unavailable, this vignette will skip deconvolution sections.

```{r xcell2-missing, eval=!has_xcell2}
cat("Package 'xCell2' is not installed; deconvolution steps are skipped.\n")
cat("Install it to run these sections:\n")
cat("  Install package 'xCell2' from Bioconductor.\n")
```

```{r run-deconvolution, eval=has_xcell2}
# First try default threshold, then relax minSharedGenes if needed.
deconv_try <- tryCatch(
  run_cell_deconvolution(
    x = vista_airway,
    method = "xCell2",
    gene_id_type = "ensembl"
  ),
  error = function(e) e
)

if (inherits(deconv_try, "error")) {
  msg <- conditionMessage(deconv_try)
  if (grepl("minSharedGenes", msg, fixed = TRUE)) {
    message("Retrying xCell2 deconvolution with minSharedGenes = 0.6")
    deconv_try <- tryCatch(
      run_cell_deconvolution(
        x = vista_airway,
        method = "xCell2",
        gene_id_type = "ensembl",
        minSharedGenes = 0.6
      ),
      error = function(e) e
    )
  }
}

if (inherits(deconv_try, "error")) {
  has_deconv <- FALSE
  message("Deconvolution could not be completed in this environment:\n", conditionMessage(deconv_try))
} else {
  vista_deconv <- deconv_try
  has_deconv <- TRUE
  vista_deconv
}
```

```{r deconvolution-skipped, eval=has_xcell2 && !has_deconv}
cat("xCell2 is installed, but deconvolution did not complete for this dataset/reference combination.\n")
cat("Downstream deconvolution plots and tables are skipped.\n")
cat("Tip: try a lower minSharedGenes or provide a different xcell2_reference.\n")
```

# Inspect Cell Fractions

```{r inspect-fractions, eval=has_xcell2 && has_deconv}
cell_fractions <- get_cell_fractions(vista_deconv)

dim(cell_fractions)
cell_fractions[1:min(4, nrow(cell_fractions)), 1:min(6, ncol(cell_fractions))]
```

# Plot Cell-Type Composition

```{r celltype-barplot, eval=has_xcell2 && has_deconv, fig.height=8}
plot_celltype_barplot(
  x = vista_deconv,
  group_column = "cond_long",
  top_n = 12,
  collapse_other = TRUE,
  normalize = "sample",
  facet_by_group = TRUE,
  font_size = 11
)
```

# Group-Level Dot Plot

This plot summarizes deconvolution signal by treatment group while keeping sample-level points visible.

```{r group-summary-plot, eval=has_xcell2 && has_deconv, fig.height=7}
get_celltype_group_dotplot(
  x = vista_deconv,
  group_column = "cond_long",
  top_n = 12,
  summary_fun = "mean",
  error = "se",
  add_points = TRUE,
  point_size = 2.5,
  font_size = 11
)
```

# Cell-Type Heatmap

The heatmap is useful to inspect sample-level deconvolution structure and concordance within groups.

```{r celltype-heatmap, eval=has_xcell2 && has_deconv, fig.height=8}
get_celltype_heatmap(
  x = vista_deconv,
  group_column = "cond_long",
  top_n = 20,
  transform = "zscore",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_values = FALSE,
  font_size = 10
)
```

# Notes on Interpretation

-   xCell2 outputs are typically **enrichment-like abundance scores** rather than absolute percentages.
-   Use group-level differences and consistency across replicates as the primary interpretation signal.
-   Treat results as hypothesis-generating; validate with orthogonal assays when possible.

# Session Info

```{r session-info}
sessionInfo()
```
