---
title: "Enrichment Chord Diagrams with VISTA"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    fig_width: 8
    fig_height: 6
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Enrichment Chord Diagrams with VISTA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# The Problem

Functional enrichment analysis (MSigDB, GO, KEGG) returns a table of
significant pathways.  A standard dot plot or bar plot answers *"what pathways
are enriched?"* -- but it hides two critical questions:

1. **Which genes drive multiple enriched pathways?**
   A handful of "hub genes" often appear in 3--5 enriched terms simultaneously.
   These genes are disproportionately important for interpreting your results,
   yet they are invisible in a standard enrichment dot plot.

2. **How redundant are my top pathways?**
   If your top 8 pathways all share the same 15 genes, they are not 8
   independent biological signals -- they are one signal reported 8 times.  A
   dot plot cannot reveal this; a chord diagram can.

## Why a chord diagram?

Enrichment results have a **many-to-many** structure: genes belong to multiple
pathways, and pathways share genes.  This is relational data -- the natural
visualization is a chord diagram where:
- **Sectors** represent pathways (on one side) and genes (on the other)
- **Chords** connect each gene to the pathways it belongs to
- **Chord colour** encodes fold-change direction or magnitude

This is fundamentally different from genome-positional plots (where a circular
layout is cosmetic) because the **connections themselves are the data**.

## What you will learn

- How to create a basic enrichment chord diagram with `get_enrichment_chord()`
- How to colour chords by pathway membership, regulation direction, or
  continuous fold-change
- How to isolate **hub genes** -- genes shared across multiple pathways
- How to control the visual output for publication-ready figures

# Setup

```{r load-packages}
library(VISTA)
library(ggplot2)
library(dplyr)
library(magrittr)
```

# Prepare Data

We use the **airway** dataset (Himes et al. 2014) -- RNA-seq from human airway
smooth muscle cells treated with dexamethasone.  The same dataset is used in
the main VISTA workflow vignette.

```{r load-airway}
library(airway)
library(org.Hs.eg.db)

data("airway", package = "airway")

# Extract counts and sample info
counts_matrix <- SummarizedExperiment::assay(airway, "counts")
sample_metadata <- as.data.frame(SummarizedExperiment::colData(airway))
sample_metadata$treatment <- ifelse(
  sample_metadata$dex == "trt", "Dexamethasone", "Untreated"
)

count_data <- as.data.frame(counts_matrix) %>% tibble::as_tibble()
count_data$gene_id <- rownames(counts_matrix)
count_data <- count_data[, c("gene_id", colnames(counts_matrix))]

sample_info <- sample_metadata %>%
  tibble::as_tibble() %>%
  dplyr::rename("sample_names" = "Run") %>%
  dplyr::mutate(sample_names = as.character(sample_names))
```

## Create VISTA object

```{r create-vista}
vista <- create_vista(
  counts       = count_data,
  sample_info  = sample_info,
  column_geneid = "gene_id",
  group_column     = "treatment",
  group_numerator  = "Dexamethasone",
  group_denominator = "Untreated",
  method       = "deseq2",
  min_counts   = 10,
  min_replicates = 2,
  log2fc_cutoff = 1.0,
  pval_cutoff   = 0.05,
  p_value_type  = "padj"
)

vista <- set_rowdata(
  vista,
  orgdb   = org.Hs.eg.db,
  columns = c("SYMBOL", "GENENAME", "ENTREZID"),
  keytype = "ENSEMBL"
)

vista
```

## Run enrichment

We need enrichment results as input.  Let's run MSigDB Hallmark and C2
canonical pathways for upregulated genes, plus Hallmark sets for downregulated
and combined (Up+Down) genes.

```{r run-enrichment, eval=requireNamespace("msigdbr", quietly=TRUE)}
comp_name <- names(comparisons(vista))[1]

# Hallmark gene sets (broad biological themes)
msig_hallmark <- get_msigdb_enrichment(
  vista,
  sample_comparison = comp_name,
  regulation  = "Up",
  msigdb_category = "H",
  species     = "Homo sapiens",
  from_type   = "ENSEMBL"
)

# C2 curated pathways (more granular, more gene overlap)
msig_c2 <- get_msigdb_enrichment(
  vista,
  sample_comparison = comp_name,
  regulation  = "Up",
  msigdb_category = "C2",
  species     = "Homo sapiens",
  from_type   = "ENSEMBL"
)

# Downregulated genes -- Hallmark
msig_down <- get_msigdb_enrichment(
  vista,
  sample_comparison = comp_name,
  regulation  = "Down",
  msigdb_category = "H",
  species     = "Homo sapiens",
  from_type   = "ENSEMBL"
)

# Combined Up+Down genes -- Hallmark
msig_both <- get_msigdb_enrichment(
  vista,
  sample_comparison = comp_name,
  regulation  = "Both",
  msigdb_category = "H",
  species     = "Homo sapiens",
  from_type   = "ENSEMBL"
)
```

Preview what standard enrichment looks like:

```{r enrichment-preview, eval=requireNamespace("msigdbr", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  head(msig_hallmark$enrich@result[, c("Description", "p.adjust", "Count")])
}
```

This table tells us *what* is enriched.  But it does not tell us whether these
pathways share the same genes or are independently driven.  That is the problem
the chord diagram solves.

# Basic Chord Diagram

## Pathway-coloured (no VISTA object needed)

The simplest form: just pass an enrichment result.  Each chord is coloured by
its source pathway, making it easy to trace which genes belong to which term.

```{r chord-basic, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_chord(
    msig_hallmark,
    top_n    = 6,
    color_by = "pathway",
    title    = "Hallmark Enrichment (Up-regulated, top 6 pathways)"
  )
}
```

**Reading the plot:**

- Bold-labelled sectors on the left/top are **pathways**
- Smaller-labelled sectors on the right/bottom are **genes**
- Each chord connects a gene to a pathway it belongs to
- Genes with chords reaching **multiple** pathway sectors are hub genes
- Wider pathway sectors contain more genes

## Understanding the output

`get_enrichment_chord()` returns data invisibly so you can inspect the results
programmatically:

```{r chord-inspect, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  result <- get_enrichment_chord(
    msig_hallmark,
    top_n    = 6,
    color_by = "pathway"
  )

  # Gene-pathway membership table
  head(result$gene_data, 10)

  # Hub genes (appearing in 2+ pathways)
  cat("\nHub genes:", paste(head(result$hub_genes, 15), collapse = ", "), "\n")
  cat("Total hub genes:", length(result$hub_genes), "\n")
}
```

# Fold-Change Colouring

The real analytical power comes when chords encode fold-change from your DE
results.  This requires passing the VISTA object and comparison name.

## Continuous fold-change gradient

```{r chord-foldchange, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_chord(
    msig_hallmark,
    vista_obj  = vista,
    comparison = comp_name,
    top_n      = 6,
    color_by   = "foldchange",
    display_id = "SYMBOL",
    title      = "Chord coloured by log2 fold-change"
  )
}
```

**What to look for:**

- **Red chords** indicate strongly upregulated genes
- **Blue/white chords** would indicate weakly changed or downregulated genes
  (rare here since we enriched on upregulated genes, but meaningful when using
  `regulation = "Both"`)
- The gradient legend (bottom-left) shows the fold-change scale

## Regulation-based colouring

A categorical alternative -- chords coloured as Up (red), Down (green), or
Other (grey):

```{r chord-regulation, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_down$enrich) && nrow(msig_down$enrich@result) > 0) {
  get_enrichment_chord(
    msig_down,
    vista_obj  = vista,
    comparison = comp_name,
    top_n      = 6,
    color_by   = "regulation",
    display_id = "SYMBOL",
    title      = "Down-regulated pathways (regulation colouring)"
  )
}
```

## Sort gene sectors by fold-change

You can order the gene side of the chord by fold-change to make the strongest
drivers appear first.

```{r chord-sorted-genes, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_chord(
    msig_hallmark,
    vista_obj     = vista,
    comparison    = comp_name,
    top_n         = 6,
    color_by      = "foldchange",
    gene_order_by = "abs_foldchange", # or "foldchange"
    display_id    = "SYMBOL",
    title         = "Gene sectors sorted by |log2FC|"
  )
}
```

This is particularly useful for enrichments run with `regulation = "Both"`,
where a pathway may contain a mix of up and downregulated genes.

## Combined Up + Down genes (fold-change coloured + FC-sorted)

This example uses an enrichment built from both up and down genes, colours
chords by fold-change, and orders gene sectors by fold-change.

```{r chord-both-fc-sorted, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_both$enrich) && nrow(msig_both$enrich@result) > 0) {
  get_enrichment_chord(
    msig_both,
    vista_obj     = vista,
    comparison    = comp_name,
    top_n         = 6,
    color_by      = "foldchange",
    gene_order_by = "foldchange",
    display_id    = "SYMBOL",
    title         = "Up + Down enrichment: fold-change coloured and FC-sorted genes"
  )
}
```

# Hub Gene Analysis

## Filtering to shared genes

The `min_pathways` parameter is the key analytical control.  Setting it to
`2` removes all genes that appear in only one pathway, revealing only the
**hub genes** that bridge multiple enriched terms.

```{r chord-hub, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  hub_result <- get_enrichment_chord(
    msig_hallmark,
    vista_obj    = vista,
    comparison   = comp_name,
    top_n        = 8,
    min_pathways = 2,
    color_by     = "foldchange",
    display_id   = "SYMBOL",
    title        = "Hub genes only (shared across 2+ pathways)"
  )

  cat("Hub genes found:", length(hub_result$hub_genes), "\n")
  cat("Genes:", paste(head(hub_result$hub_genes, 20), collapse = ", "), "\n")
}
```

**Interpretation:**

- Every gene shown has chords reaching at least 2 pathway sectors
- These genes are disproportionately responsible for the enrichment signal
- If the same 10 genes drive all 8 pathways, the enrichment is one
  signal, not eight

## Increasing the threshold

For dense enrichments (e.g. C2 canonical pathways), you can raise the
threshold further:

```{r chord-hub-strict, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_c2$enrich) && nrow(msig_c2$enrich@result) > 0) {
  c2_result <- get_enrichment_chord(
    msig_c2,
    vista_obj    = vista,
    comparison   = comp_name,
    top_n        = 8,
    min_pathways = 3,
    max_genes    = 30,
    color_by     = "regulation",
    display_id   = "SYMBOL",
    title        = "C2 Pathways: genes in 3+ pathways"
  )

  if (length(c2_result$hub_genes)) {
    cat("Top hub genes:", paste(head(c2_result$hub_genes, 15), collapse = ", "), "\n")
  }
}
```

# Visual Customisation

## Custom pathway colours

Pass a named colour vector to override the default HCL palette:

```{r chord-custom-colors, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  # Get the top pathway names first
  top_pathways <- head(msig_hallmark$enrich@result$Description, 5)

  # Define custom colors
  custom_cols <- c(
    "#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51"
  )
  names(custom_cols) <- top_pathways

  get_enrichment_chord(
    msig_hallmark,
    top_n          = 5,
    color_by       = "pathway",
    pathway_colors = custom_cols,
    title          = "Custom pathway colours"
  )
}
```

## Adjusting layout

Fine-tune gap size, label size, and transparency:

```{r chord-layout, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_chord(
    msig_hallmark,
    top_n        = 5,
    color_by     = "pathway",
    gap_degree   = 4,          # wider gaps between sectors
    label_cex    = 0.6,        # smaller labels
    transparency = 0.2,        # more opaque chords
    title        = "Adjusted layout"
  )
}
```

## Selecting specific pathways

Instead of top N by p-value, pick pathways by name:

```{r chord-specific, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  available <- msig_hallmark$enrich@result$Description
  # Pick specific pathways of interest (if available)
  pick <- intersect(
    c("HALLMARK_P53_PATHWAY",
      "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
      "HALLMARK_HYPOXIA",
      "HALLMARK_APOPTOSIS"),
    available
  )

  if (length(pick) >= 2) {
    get_enrichment_chord(
      msig_hallmark,
      pathways   = pick,
      vista_obj  = vista,
      comparison = comp_name,
      color_by   = "foldchange",
      display_id = "SYMBOL",
      title      = "Selected pathways of interest"
    )
  }
}
```

# Comparison with Standard Visualisations

To appreciate what the chord diagram adds, compare it with the existing
VISTA enrichment views on the same data:

## Dot plot: *what* is enriched

```{r compare-dotplot, fig.width=9, fig.height=6, eval=requireNamespace("msigdbr", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_plot(msig_hallmark$enrich, top_n = 8)
}
```

The dot plot answers: *which pathways are significant?*  It shows p-values and
gene counts but does not reveal which specific genes are shared.

## Pathway heatmap: *expression* of pathway genes

```{r compare-heatmap, fig.width=8, fig.height=8, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("ComplexHeatmap", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_pathway_heatmap(
    vista_obj  = vista,
    enrichment = msig_hallmark$enrich,
    samples    = c("Untreated", "Dexamethasone"),
    top_n      = 3,
    gene_mode  = "union",
    max_genes  = 40,
    value_transform    = "zscore",
    display_id         = "SYMBOL",
    annotate_columns   = TRUE,
    summarise_replicates = FALSE,
    show_row_names     = TRUE
  )
}
```

The heatmap answers: *how do pathway genes behave across samples?*  It shows
expression patterns but does not reveal gene sharing across pathways.

## Chord diagram: *topology* of gene--pathway relationships

```{r compare-chord, fig.width=9, fig.height=9, eval=requireNamespace("msigdbr", quietly=TRUE) && requireNamespace("circlize", quietly=TRUE)}
if (!is.null(msig_hallmark$enrich) && nrow(msig_hallmark$enrich@result) > 0) {
  get_enrichment_chord(
    msig_hallmark,
    vista_obj    = vista,
    comparison   = comp_name,
    top_n        = 8,
    min_pathways = 2,
    color_by     = "foldchange",
    display_id   = "SYMBOL",
    title        = "Hub genes bridging enriched pathways"
  )
}
```

The chord diagram answers: *which genes bridge which pathways, and are they
going up or down?*

| View                  | Question it answers                        | Key insight                        |
|-----------------------|--------------------------------------------|------------------------------------|
| `get_enrichment_plot` | What pathways are significant?             | Statistical significance and size  |
| `get_pathway_heatmap` | How do pathway genes behave across samples?| Expression patterns and clustering |
| `get_enrichment_chord`| Which genes connect which pathways?        | Hub genes and pathway redundancy   |

These three views are **complementary**, not competing.  Together they form a
complete picture of functional enrichment.

# Practical Guidelines

## When to use chord diagrams

- After running enrichment to understand **redundancy** in your top terms
- When you need to identify **hub genes** for follow-up validation (qPCR,
  western blot)
- When presenting to collaborators who need a visual map of gene--pathway
  relationships

## When NOT to use chord diagrams

- With > 10 pathways and > 50 genes (becomes unreadable -- use `max_genes` cap)
- As a replacement for dot plots (you still need the dot plot for significance)
- For genome-positional data (use `get_foldchange_chromosome_plot()` instead)

## Recommended defaults for common scenarios

| Scenario                    | `top_n` | `min_pathways` | `color_by`    |
|-----------------------------|---------|----------------|---------------|
| First exploration           | 6       | 1              | `"pathway"`   |
| Identify hub genes          | 8       | 2              | `"foldchange"`|
| Publication figure          | 5--6    | 2              | `"foldchange"`|
| Present to collaborators    | 4--5    | 1              | `"regulation"`|
| Dense pathway sets (C2/C5)  | 6       | 3              | `"regulation"`|

# Session Information

```{r session-info}
sessionInfo()
```

# References

- Himes BE et al. (2014). "RNA-Seq transcriptome profiling identifies
  CRISPLD2 as a glucocorticoid responsive gene that modulates cytokine
  function in airway smooth muscle cells." *PLoS One* 9(6): e99625.
- Gu Z et al. (2014). "circlize implements and enhances circular
  visualization in R." *Bioinformatics*, 30(19), 2811--2812.
- Wu T et al. (2021). "clusterProfiler 4.0: A universal enrichment tool for
  interpreting omics data." *The Innovation*, 2(3), 100141.
- Subramanian A et al. (2005). "Gene set enrichment analysis: a
  knowledge-based approach for interpreting genome-wide expression profiles."
  *PNAS*, 102(43), 15545--15550.
