---
title: "Raincloud Plots in VISTA"
author: "VISTA Development Team"
package: VISTA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Raincloud Plots in VISTA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

has_ggrain <- requireNamespace("ggrain", quietly = TRUE)
```

# Overview

Raincloud plots combine three visual layers:

1. distribution shape (half-violin),
2. robust summary (boxplot), and
3. individual observations (jittered points).

This vignette shows raincloud plotting for both expression and fold-change data
in VISTA, including line-connection control with `id.long.var` and optional
statistical annotations.

# Create a VISTA object

```{r create-vista}
library(VISTA)
library(ggplot2)

data("count_data", package = "VISTA")
data("sample_metadata", package = "VISTA")

# Keep runtime modest for vignette rendering
count_small <- count_data[1:1000, ]

vista <- create_vista(
  counts = count_small,
  sample_info = sample_metadata,
  column_geneid = "gene_id",
  group_column = "cond_long",
  group_numerator = "treatment1",
  group_denominator = "control",
  method = "deseq2",
  min_counts = 10,
  min_replicates = 1
)

comp_names <- names(comparisons(vista))
top_up <- get_genes_by_regulation(
  vista,
  sample_comparisons = comp_names[1],
  regulation = "Up"
)
n_select_genes = 50
selected_genes <- stats::na.omit(utils::head(top_up$gene_id, n_select_genes))
if (!length(selected_genes)) {
  selected_genes <- rownames(vista)[1:n_select_genes]
}
```

```{r ggrain-missing, eval=!has_ggrain}
cat("Package 'ggrain' is not installed; raincloud examples are skipped.\n")
```

# Expression Raincloud

## Basic expression raincloud

```{r expr-rain-basic, fig.width=9, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE
)
```

## `summarise = FALSE` vs `summarise = TRUE`

Use `summarise = FALSE` to show per-sample distributions, and
`summarise = TRUE` to aggregate replicate-level values by group.

```{r expr-rain-nosummarise, fig.width=10, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = FALSE,
  facet = FALSE,
  id.long.var = "gene"
)
```

```{r expr-rain-summarise, fig.width=10, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE,
  id.long.var = "gene"
)
```

## Expression raincloud with lines and p-values

```{r expr-rain-lines-stats, fig.width=9, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE,
  id.long.var = "gene",
  stats_group = TRUE,
  stats_method = "wilcox.test",
  p.label = "p.format"
)
```

## Label dots by gene ID (`facet = FALSE`)

```{r expr-rain-labels, fig.width=10, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE,
  label_points = TRUE,
  label_column = "gene",
  label_size = 3
)
```

If your object has symbol annotations in `rowData(vista)` (or you provide
`display_from`/`display_orgdb`), you can label directly in symbol space:

```{r expr-rain-symbol-demo, eval=FALSE}
get_expression_raincloud(
  vista,
  genes = c("NFKBIA", "KLF6", "PER1"),
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE,
  label_points = TRUE,
  display_id = "SYMBOL"
)
```

When `facet = TRUE`, labels can still be added, but readability usually drops
quickly as the number of genes/points increases. Keep the gene set small.

```{r expr-rain-labels-facet, fig.width=10, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes[1:2],
  value_transform = "log2",
  summarise = TRUE,
  facet = TRUE,
  label_points = TRUE,
  label_column = "gene",
  label_size = 2.8
)
```

## Flipped expression raincloud

Raincloud plots can be visually emphasized in a horizontal layout by combining a
left-side raincloud with `coord_flip()`.

```{r expr-rain-flipped, fig.width=10, fig.height=6, eval=has_ggrain}
get_expression_raincloud(
  vista,
  genes = selected_genes,
  value_transform = "log2",
  summarise = TRUE,
  facet = FALSE,
  rain_side = "r",
  id.long.var = "gene"
) +
  ggplot2::coord_flip()
```

# Fold-Change Raincloud

## Basic fold-change raincloud

```{r fc-rain-basic, fig.width=9, fig.height=6, eval=has_ggrain}
get_foldchange_raincloud(
  vista,
  sample_comparisons = comp_names,
  facet = TRUE
)
```

## Fold-change raincloud with gene trajectories and p-values

```{r fc-rain-lines-stats, fig.width=9, fig.height=6, eval=has_ggrain}
get_foldchange_raincloud(
  vista,
  sample_comparisons = comp_names,
  facet = FALSE,
  id.long.var = "gene_id",
  stats_group = TRUE,
  stats_method = "t.test"
)
```

## Label dots by gene ID for fold-change raincloud

```{r fc-rain-labels, fig.width=10, fig.height=6, eval=has_ggrain}
get_foldchange_raincloud(
  vista,
  sample_comparisons = comp_names,
  facet = FALSE,
  label_points = TRUE,
  label_column = "gene_id",
  label_size = 2.8
)
```

```{r fc-rain-symbol-demo, eval=FALSE}
get_foldchange_raincloud(
  vista,
  genes = c("NFKBIA", "KLF6", "PER1"),
  sample_comparisons = comp_names,
  facet = FALSE,
  label_points = TRUE,
  display_id = "SYMBOL"
)
```

## Flipped fold-change raincloud

```{r fc-rain-flipped, fig.width=10, fig.height=6, eval=has_ggrain}
get_foldchange_raincloud(
  vista,
  sample_comparisons = comp_names,
  facet = FALSE,
  rain_side = "r",
  id.long.var = "gene_id"
) +
  ggplot2::coord_flip()
```

# Why this is harder outside VISTA

Outside VISTA, producing equivalent raincloud plots is more involved because you
must manually:

1. extract and harmonize DE tables per comparison,
2. reshape to long format for plotting,
3. track grouping and palette consistency,
4. map repeated-measure identifiers for line connections, and
5. add and control statistical annotations per plotting context.

A minimal non-VISTA workflow typically requires custom wrangling and multiple
plot-specific settings:

```{r outside-vista-sketch, eval=FALSE}
# 1) Build long expression/fold-change tables manually
# 2) Join sample metadata and comparison metadata
# 3) Validate IDs for repeated measures (id.long.var)
# 4) Create raincloud layers and palette mapping
# 5) Add statistical comparisons and label formatting
# 6) Repeat the process for each analysis object/comparison set
```

In VISTA, these steps are encapsulated in `get_expression_raincloud()` and
`get_foldchange_raincloud()` while staying consistent with the rest of the
plotting API.

## Session information

```{r session-info}
sessionInfo()
```
